Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of TESS

### 2023-08-11 paper/ms.tex
update ms

@@

\author{-----SUGGESTED CONTRIBUTORS, PENDING DISCUSSION, COMMENTS, ANALYSES, ETC-----}
%
\author[0000-0001-6381-515X]{Luisa M. Rebull}
\affiliation{\caltech}

@@

% 250 word max!
\begin{abstract}
  Complex quasiperiodic variables (CQVs) are low-mass
  pre-main-sequence stars with nearly periodic optical modulation.
  The modulation is likely induced by dust or gas clumps in orbit at
  the Keplerian corotation radius.  Here, we report new CQVs
  discovered in TESS short-cadence data collected between July~2018
  and Sep~2022.  Our search of \nstarssearched\ K and M dwarfs with
  $T$$<$16 and $d$$<$150\,pc yielded \ngoods\ high-quality CQVs.  Most
  of these discoveries are new, and they include the brightest
  ($T$$\approx$9.5) and closest ($d$$\approx$20\,pc) examples of this
  class of object known, as well as the most massive
  ($\approx$0.6-0.8\,$M_\odot$) and the oldest ($\approx$200\,Myr).  A few
  objects are outliers among outliers; LP 12-502 for instance shows a
  recurring ``dip complex'' with a period and total duration that are
  fixed over more than 1{,}000 cycles, but which in detail contains
  anywhere from four to eight local minima per cycle, and at times
  displays multiple periods simultaneously.  Broadly
  speaking, we find that none of the CQVs maintain a fixed light curve
  shape over timescales of more than a few hundred cycles, and we
  revisit the arguments for why transient corotating material is the
  most viable explanation.  We expect that our sample will facilitate
  future efforts aimed at connecting these objects to the broader
  contexts of star, disk, and perhaps even exoplanet evolution. 
  %For instance, if the transiting material is dust, then the rate of
  %dip evolution implies that of order an asteroid belt's worth of mass
  %passes through these structures over their lifetime of
  %$\approx$$10^8$ years.
\end{abstract}

@@

Once a star's period $P$ was identified, we binned the phased light
curve to 100 points per cycle.  To separate ``sharp'' local minima
from smooth spot-induced variability, we then iteratively fit robust
penalized splines to the wrapped phase-folded light curve, excluding
points more than two standard deviations away from the local continuum
\citep{2019AJ....158..143H}.  The maximum number of equidistant spline
knots per cycle is the parameter in this framework that controlled the
meaning of ``sharp'' --- we allowed at most 10 such knots per cycle,
though for most stars fewer knots were preferred based on
cross-validation using an $\ell^2$-norm penalty.  An example fit is
shown in panel {\it (e)} of Figure~\ref{fig:vetting}.

@@

In this work, we searched 120-second cadence TESS data collected
between July~2018 and Sep~2022 for complex quasiperiodic variables
(CQVs).  Our search sample included \nstarssearched\ K and M dwarfs
within 150\,pc, yielding a selection function with $\gtrsim 80\%$ completeness within 30\,pc,
and $\lesssim 10\%$ completeness at distances exceeding 100\,pc.

@@

\begin{enumerate}[leftmargin=*]
	%
	\item CQVs are quasiperiodic.  The mean periods remain fixed over the
		$>$1{,}000-cycle baseline of available observations; but the light curve shapes
		always evolve (Figure~\ref{fig:evoln}).
  %
  \item The same CQV can show dips with similar but clearly
    distinct periods.  LP 12-502, for instance, showed dips with four
    distinct periods within $\pm 0.3\%$ of its fundamental period,
    sometimes simultaneously, and each lasting for up to 50 cycles
    (Figure~\ref{fig:lpriver0}).
  %
    \item CQVs evolve over timescales that are both secular
    ($>$100\,cycles) and impulsive ($<$1 cycle).  ``State-switches''
    can cause dips to collapse instantaneously, and are often but not
    exclusively linked with observed optical flares.  Dip growth
    however seems to happen over durations of at least ten cycles, and
    slow dip decay can also occur.
  %
    \item The rate of dip evolution can be used to place a model-dependent
    constraint on amount of material that is either being accreted or ejected
    during the state changes (Section~\ref{subsec:massflux}).
    Order of magnitude estimates require at least an asteroid belt's worth
    of dust ($10^{-4}$\,$M_\oplus$) over $10^8$ years, or 100 times more 
    material if the occulting matter is gas.
  %
  \item The CQV phenomenon persists for $\gtrsim$150\,Myr, based on
    the existence of multiple CQVs in AB Dor, the Pleiades, and
    Psc-Eri (Section~\ref{subsec:starprops}).  It may even extend to
    200\,Myr, based on the one CQV we found in the Carina Near
    moving group (TIC~294328887; $\approx$200\,Myr).  The lack of
    detected CQVs in the Hyades and Praesepe suggests that the
    lifetime of the phenomenon is likely limited to the first few
    hundred million years.
	%
    \item The duty cycle for CQVs seems to be $\approx$$75$\%, based on
    the fraction of bonafide CQVs that turned on or turned off during
    TESS re-observations, two years after their initial observation.
  %
    \item While the majority of detected CQVs have masses of
    0.1-0.5\,$M_\odot$, the upper mass limit for the CQV phenomenon
    might extend into the K dwarf regime ($<$$0.8$\,$M_\odot$).
    TIC~405754448 and TIC~405910546, with masses of $\approx$0.8 and
    $\approx$0.6 solar masses, are the two massive outliers
    (Section~\ref{subsec:extreme}), and we encourage additional
    scrutiny of these objects in future work.
  %
    \item Surprising analogs to the M  and K dwarf CQVs exist in
    magnetic B stars, but only those with multipolar field topologies.
    This suggests that the CQV dips and warps are strongly sculpted
    by the stellar magnetic fields, which are probably also multipolar.
  %
    \item The closest CQVs to the Sun are at distances of 15-20\,pc, and
    the brightest CQVs have $V$$\approx$12 ($J$$\approx$7.5).  This
    conclusion is an extrapolation from the CQV sample derived in this work
    being $\gtrsim$80\% complete within 20\,pc.  Expanding our analysis of
    the TESS data to the full frame images would yield a truly
    volume-limited selection function, and would expand the CQV census
    by about a factor of two within 50\,pc, and by a factor of ten
    within 100\,pc.
\end{enumerate}

@@

The distinction between gas or dust is important because it will
clarify whether the CQV phenomenon is intrinsic, so that the material
comes from the star, or extrinsic, so that it is sourced through some
generic evolutionary phase of circumstellar disks.  This knowledge would in
turn propagate to our understanding of whether observing the
phenomenon is primarily teaching us about dust production and
processing in circumstellar disks, or whether it is teaching us about
the ability of cold gas to remain stable in hot stellar coronae for
long durations.
Observationally, acquisition of medium- or high-resolution time-series
spectroscopy holds the best chance at resolving the gas vs.~dust
question.  Given the observed evolutionary timescales, such data
should be acquired simultaneously with photometric time-series
observations to help clarify its interpretation.

In both the gas and dust scenarios, CQVs are preferentially viewed
edge-on.  This implies that after correcting for the line-of-sight
inclination, at least one third of low mass stars
\citep{2022AJ....163..144G} trap circumstellar material in the same
way.  This suggests that CQVs should preferentially show edge-on
transiting planets, at larger distances from the star than the
corotating material.  Given these points, important follow-up
observations should include searching for outer transiting planets,
and measuring equatorial velocities in order to test whether the
stellar inclination angles are indeed preferentially edge-on.

>>>>

\author[0000-0001-6381-515X]{Luisa M. Rebull}
\affiliation{\caltech}

@@

\author{-----SUGGESTED CONTRIBUTORS, PENDING DISCUSSION, COMMENTS, ANALYSES, ETC-----}
%

@@

% 250 word max!
\begin{abstract}
Complex quasiperiodic variables (CQVs) are low-mass pre-main-sequence
stars with nearly periodic optical modulation.  The modulation is
likely induced by dust or gas clumps in orbit at the Keplerian
corotation radius.  Here, we report new CQVs discovered in TESS
short-cadence data collected between July~2018 and Sep~2022.  Our
search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
$d$$<$150\,pc yielded \ngoods\ high-quality CQVs.  Most of these
discoveries are new, and they include the brightest
($T$$\approx$9.5), closest ($d$$\approx$20\,pc), most massive
($\approx$0.6-0.8\,$M_\odot$), and the oldest ($\approx$200\,Myr)
examples of this class of object currently known.  A few objects are
outliers: LP 12-502 for instance showed a ``dip complex'' with a
period and duration that were fixed over 1,500 cycles, but in
detail, this system exhibited between four and eight local minima
per cycle, and at times also displayed multiple periods
simultaneously.  We demonstrate that transient corotating material
is the most viable explanation for this class of object, and present
evidence supporting the hypothesis that this material is sculpted by
stellar magnetic fields dominated by high order multipoles.  We
expect that our sample will facilitate future efforts aimed at
connecting CQVs to the broader contexts of star, disk, and perhaps
even exoplanet evolution. 	
\end{abstract}

@@

Once a star's period $P$ was identified, we binned the phased light
curve to 100 points per cycle.  To separate ``sharp'' local minima
from smooth spot-induced variability, we then iteratively fit robust
penalized splines to the wrapped phase-folded light curve, excluding
points more than two standard deviations away from the local continuum
\citep{2019AJ....158..143H}.  The maximum number of equidistant spline
knots per cycle is the parameter in this framework that controlled the
meaning of ``sharp'' --- we allowed at most 10 such knots per cycle,
though for most stars fewer knots were preferred based on
cross-validation using an $\ell^2$-norm penalty.  An example fit is
shown in panel {\it (e)} of Figure~\ref{fig:vet}.

@@

In this work, we searched 120-second cadence TESS data collected
between July~2018 and Sep~2022 for complex quasiperiodic variables
(CQVs).  The target stars were \nstarssearched\ K and M dwarfs
within 150\,pc.  The selection function was $\gtrsim 80\%$ complete within 30\,pc,
and $\lesssim 10\%$ complete at distances exceeding 100\,pc.

@@

\begin{enumerate}[leftmargin=*]
	%
	\item CQVs are quasiperiodic.  The mean periods remain fixed over the
		$>$1{,}000-cycle baseline of available observations; but the light curve shapes
		always evolve (Figure~\ref{fig:evoln}).
  %
  \item The same CQV can show multiple periods simultaneously.
     LP 12-502, for instance, showed dips with four
    distinct periods within $\pm 0.3\%$ of its fundamental period,
    sometimes simultaneously, and each lasting for up to 50 cycles
    (Figure~\ref{fig:lpriver0}).
  %
    \item CQVs evolve over timescales that are both secular
    ($>$100\,cycles) and impulsive ($<$1 cycle).  ``State-switches''
    can cause dips to collapse instantaneously, and are often but not
    exclusively linked with observed optical flares.  Dip growth
    however seems to happen over durations of at least ten cycles, and
    slow dip decay can also occur.
  %
    \item The rate of dip evolution can be used to place a model-dependent
    constraint on how much material is either being accreted or ejected
    during the state changes (Section~\ref{subsec:massflux}).
    Order of magnitude estimates require at least an asteroid belt's worth
    of dust ($10^{-4}$\,$M_\oplus$) over $10^8$ years, or 100 times more 
    material if the occulting matter is gas.
  %
  \item The CQV phenomenon persists for $\gtrsim$150\,Myr, based on
    the existence of multiple CQVs in AB Dor, the Pleiades, and
    Psc-Eri (Section~\ref{subsec:starprops}).  It may even extend to
    200\,Myr, based on the one CQV we found in the Carina Near
    moving group (TIC~294328887; $\approx$200\,Myr).  The lack of
    detected CQVs in the Hyades and Praesepe suggests that the
    lifetime of the phenomenon is likely limited to the first few
    hundred million years.
	%
    \item The duty cycle for CQVs seems to be $\approx$$75$\%, based on
    the fraction of bonafide CQVs that either turned on or turned off during
    TESS re-observations, two years after the initial observation.
  %
    \item While the majority of detected CQVs have masses of
    0.1-0.5\,$M_\odot$, the upper mass limit for the CQV phenomenon
    might extend into the K dwarf regime ($<$$0.8$\,$M_\odot$).
    TIC~405754448 and TIC~405910546, with masses of $\approx$0.8 and
    $\approx$0.6\,$M_\odot$, are the two massive outliers
    (Section~\ref{subsec:extreme}), and we encourage additional
    scrutiny of these objects in future work.
  %
    \item Surprising analogs to the M  and K dwarf CQVs exist in
    magnetic B stars, but only those with multipolar field topologies.
    Since most magnetic B stars have dipolar magnetic fields,
    this suggests that the CQV dips and warps are similarly being sculpted
    by the stellar magnetic fields, and that the magnetic
    fields themselves are probably also multipolar.
  %
    \item The closest CQVs to the Sun are at distances of 15-20\,pc, and
    the brightest have $V$$\approx$12 ($J$$\approx$7.5).  This
    conclusion is an extrapolation from the CQV sample derived in this work
    being $\gtrsim$80\% complete within 30\,pc.  Expanding our analysis of
    the TESS data to the full frame images would yield a truly
    volume-limited selection function, and would expand the CQV census
    by about a factor of two within 50\,pc, and by a factor of ten
    within 100\,pc.
\end{enumerate}

@@

The distinction between gas or dust is important because it will
clarify whether the CQV phenomenon is intrinsic, so that material
comes from the star, or extrinsic, so that it is sourced through some
generic evolutionary phase of circumstellar disks.
This knowledge would in
turn propagate to our understanding of whether the
phenomenon is primarily teaching us about dust production and
processing in circumstellar disks, or whether it is teaching us about
the ability of cold gas to remain stable in hot stellar coronae for
long durations.
Observationally, acquisition of medium- or high-resolution time-series
spectra holds the best chance at resolving the gas vs.~dust
question.  Given the observed evolutionary timescales, such data
should be acquired simultaneously with photometric time-series
observations to help clarify its interpretation.

%TODO: nix this paragraph?  This edge on hypothesis only holds
%TODO if the magnetic field is aligned with the stellar spin axis.
{\bf todo: nix paragraph? technically this only works if B is aligned
with stellar spin axis... which would be aligned w planetary system}
In both the gas and dust scenarios, CQVs are preferentially viewed
edge-on.  This implies that after correcting for the line-of-sight
inclination, at least one third of low mass stars
\citep{2022AJ....163..144G} trap circumstellar material in the same
way.  This suggests that CQVs should preferentially show edge-on
transiting planets, at larger distances from the star than the
corotating material.  Given these points, important follow-up
observations should include searching for outer transiting planets,
and measuring equatorial velocities in order to test whether the
stellar inclination angles are indeed preferentially edge-on.
